Annotated example of index standardization with VAST

### 2021-11-24 doc/appendix-pcod-VAST.Rmd
Tidy VAST example

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(TMB)               
library(VAST)
library(sp)
library(sdmTMB)
library(magick)
```

@@

```{r, echo = TRUE}
FieldConfig = matrix(c("IID","IID","IID","IID","IID","IID"), ncol=2, nrow=3, 
                     dimnames = list(c("Omega","Epsilon","Beta"),
                                     c("Component_1","Component_2")))
RhoConfig = c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
```

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).  Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid"=input_grid` in `fit_model`.

```{r settings, echo = TRUE}
# get coordinates in geographic coordinates from utm projection, after changing units to m
qcs_grid_ll <- qcs_grid
qcs_grid_ll$Y <- qcs_grid_ll$Y * 1000
qcs_grid_ll$X <- qcs_grid_ll$X * 1000

# with sp
coordinates(qcs_grid_ll) = ~X + Y
proj4string(qcs_grid_ll) <- CRS("+proj=utm +zone=9")
qcs_grid_ll <- as.data.frame(spTransform(qcs_grid_ll, CRS("+proj=longlat +datum=WGS84")))

## or with sf
# qcs_grid_ll <- st_as_sf(x = qcs_grid_ll,
#                         coords = c("X", "Y"),
#                         crs = "+proj=utm +zone=9")
# qcs_grid_ll <- st_transform(qcs_grid_ll, crs = "+proj=longlat +datum=WGS84")

# remove replicate locations for each year and format for VAST
qcs_grid_ll <- dplyr::filter(qcs_grid_ll, year == min(qcs_grid_ll$year))
input_grid <- cbind(Lat = qcs_grid_ll$Y, Lon = qcs_grid_ll$X, Area_km2 = 4)

settings <- make_settings(n_x = 205,
                          Region = "User", 
                          purpose = "index2", 
                          fine_scale = TRUE, 
                          zone = 9,
                          FieldConfig = FieldConfig,
                          RhoConfig = RhoConfig,
                          ObsModel = c(2,0), # this is for conventional logit-linked delta-gamma; c(10,2) for Tweedie
                          bias.correct = FALSE,
                          use_anisotropy = FALSE,
                          max_cells = Inf, # or set higher or lower to predict to greater/fewer locations
                          knot_method = "grid") # or "samples"
```

@@

```{r fit, echo = TRUE, results = 'hide'}
dir.create(paste0(here::here("doc", "appendix-VAST"))) # create folder for saved output

pcod$effort = 1 # effort is 1 when using CPUE instead of observed weight as the response

fit <- fit_model(settings = settings, 
                 Lat_i = pcod[,'lat'], 
                 Lon_i = pcod[,'lon'], 
                 t_i = pcod[,'year'], 
                 b_i = pcod[,'density'], 
                 a_i = pcod[,'effort'], 
                 input_grid = input_grid,
                 working_dir = paste0(here::here("doc", "appendix-VAST"), "/"))

```

@@

```{r save-plots, echo = TRUE, warning = FALSE}
plot(fit, 
     check_residuals = FALSE,
     working_dir = paste0(here::here("doc", "appendix-VAST"), "/"))
```

@@

```{r plot-knots, echo = TRUE}
knots <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "Data_and_knots.png"))
plot(knots)
```

@@

```{r plot-density, echo = TRUE}
density <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "ln_density-predicted.png"))
plot(density)
```

@@

```{r plot-index, echo = TRUE}
index <- magick::image_read(paste0(here::here("doc", "appendix-VAST"), "/", "Index-Biomass.png"))
plot(index)
```

>>>>

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(TMB)
library(VAST)
library(sp)
library(sdmTMB)
library(here)
```

@@

```{r, echo = TRUE}
FieldConfig <- matrix(c("IID", "IID", "IID", "IID", "IID", "IID"),
  ncol = 2, nrow = 3,
  dimnames = list(
    c("Omega", "Epsilon", "Beta"),
    c("Component_1", "Component_2")
  )
)
RhoConfig <- c("Beta1" = 0, "Beta2" = 0, "Epsilon1" = 0, "Epsilon2" = 0)
```

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).  Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.

```{r settings, echo = TRUE}
# get coordinates in geographic coordinates from UTM projection, after changing units to m
qcs_grid_ll <- qcs_grid
qcs_grid_ll$Y <- qcs_grid_ll$Y * 1000
qcs_grid_ll$X <- qcs_grid_ll$X * 1000

# with sp:
coordinates(qcs_grid_ll) <- ~ X + Y
proj4string(qcs_grid_ll) <- CRS("+proj=utm +zone=9")
qcs_grid_ll <- as.data.frame(spTransform(qcs_grid_ll, CRS("+proj=longlat +datum=WGS84")))

# or with sf:
# qcs_grid_ll <- sf::st_as_sf(
#   x = qcs_grid_ll,
#   coords = c("X", "Y"),
#   crs = "+proj=utm +zone=9"
# )
# qcs_grid_ll <- sf::st_transform(qcs_grid_ll, crs = "+proj=longlat +datum=WGS84")

# remove replicate locations for each year and format for VAST
qcs_grid_ll <- subset(qcs_grid_ll, year == min(qcs_grid_ll$year))
input_grid <- cbind(Lat = qcs_grid_ll$Y, Lon = qcs_grid_ll$X, Area_km2 = 4)

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # this is for conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

```{r fit, echo = TRUE, results = 'hide'}
 # create folder for saved output:
dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- as.numeric(1)

fit <- fit_model(
  settings = settings,
  Lat_i = pcod[, "lat"],
  Lon_i = pcod[, "lon"],
  t_i = pcod[, "year"],
  b_i = pcod[, "density"],
  a_i = pcod[, "effort"],
  input_grid = input_grid,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

@@

```{r save-plots, echo = TRUE, warning = FALSE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

@@

```{r plot-knots, echo = TRUE}
knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE}
knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
```

@@

```{r plot-index, echo = TRUE}
if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
}
if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
}
```
